---
title: Metric space change point detection
subtitle: #Do you have a subtitle? If so, write it here
titlerunning: Metric change point detection
authorrunning: #Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.

authors: 
- name: David Letscher
  address: Department of Computer Science, Saint Louis University
  email: david.letscher@slu.edu
  
- name: Darrin Speegle
  address: Department of Mathematics and Statistics, Saint Louis University
  email: darrin.speegle@slu.edu

keywords:
- change point detection
- metric space
- persistence
- tda

preamble: 
  \usepackage{amsfonts}
  \usepackage{algorithm}
  \usepackage[noend]{algpseudocode}
  \makeatletter
  \def\BState{\State\hskip-\ALG@thistlm}
  \makeatother

#PACS: 
#- PAC1
#- superPAC
    
MSC:
- MSC code 1
- MSC code 2    

abstract: |
  Let $(x_t)_{t = 1}^N$ be a time series with values in a metric space $X$, which is locally isometric to Euclidean space. A transformation of the data is proposed which produces a multi-dimensional time series of real numbers. If the original sequence of data points has a single change point in mean at time $t_0$, then with high probability the transformed data will also have a single change point in mean at time $t_0$. Applications to time series of persistence diagrams are considered.

bibliography: bibliography.bib
output: rticles::springer_article
---


# Introduction {#intro}

A simple change point detection problem is the following: suppose that $(x_t)_{t = 1}^{t_0}$ are iid Gaussian random variables with mean $\mu_0$ and variance $\sigma$, and $(x_t)_{t = t_0 + 1}^N$ are iid Gaussian with mean $\mu_1$ and variance $\sigma$. The goal is to determine whether $\mu_0 = \mu_1$, and if not, then what the value of $t_0$ is. 

```{r, echo = FALSE, fig.height = 3, fig.width = 3, fig.align = "center", message = FALSE}
set.seed(1)
dat <- data.frame(t = 1:64, y = c(rnorm(30), rnorm(34, 2, 1)))
library(ggplot2)
ggplot(dat, aes(x = t, y = y)) +
  geom_point(size = 0.5) +
  geom_vline(xintercept = 30.5, color = "red") +
  ggtitle("Change Point at Time t = 30")

```



More general problems include having multiple change points, noise that is not Gaussian, underlying signals that are not constant, and multi-dimensional signals. 

In this paper, we consider the following set-up. Let $X$ be a metric space. Supose that $f:[0,1]\to X$ is a piecewise continuous function with at most one discontinuity.  Let $(a_t)_{t = 1}^N$ be an increasing sequence of numbers in $(0, 1)$ and $\delta > 0$ such that for each $y\in [0,1]$, $B_\delta(f(y))$ is isometric to Euclidean space. Let $x_t  = f(a_t) + \epsilon_t$, where $\epsilon_t$ is a uniformly distributed random variable on $B_\delta(f(a_t))$.  Our problem is to determine whether there exists a point of discontinuity $y_0$ of $f$ and a $k < N$ such that the discontinuity is contained between $a_k$ and $a_{k +1}$. If there does exist such a $k$, then we should also estimate it.

Our motivation is that we wish to find change points in time series of **persistence diagrams**. Persistence diagrams are a way of measuring the topological structure of a point cloud in Euclidean space. More here.

# Metric space change point detection algorithm {#sec:1}

With the notation as set up in Section \ref{intro}, we proceed to describe the algorithm for change point detection for time series data in metric spaces. The main step is transforming the data into a multi-dimensional time series of real numbers in such a way as to preserve the change point, see Algorithm \ref{euclid}.

\begin{algorithm}
\caption{Transform to Real}\label{euclid}
\begin{algorithmic}[1]
\Procedure{MyProcedure}{}
\State $\textit{time\_length} \gets \text{length of }\textit{time\_series}$
\State $i \gets 1$
\While {$\textit{i < time\_length}$}
\State $\textit{A, B} \gets \text{sample(N, 2, replace = FALSE)}$
\For {$j \gets 1:time\_length$}
\State $\textit{dists[j,i]} \gets (d(A, x_j)^2 - d(B, x_j)^2)/d(A, B)$
\EndFor
\State $\textit{i++}$
\EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

Once the metric space valued time series is transformed into a multi-dimensional real valued time series, standard techniques can be used to determine if and where a change point occurs. In this paper, we use the \texttt{cpbaywave} package, as it finds change points in high dimensional, smooth data with a single discontinuity. We illustrate the usage with some examples.

\begin{example}
Suppose that the time series lives in $\mathbb {R}^M$ for some $M$. For example, we could have a time series of length 128 in 100 dimensional data, with a change point at time $t = 80$. We consider two examples in this case; in both examples, the data has mean zero in all dimensions until time 80. In the first case, it has mean 0.1 in all dimensions from dimension 81 through 128 and in the second case it has mean 1 and all dimensions from dimension 81 through 128.

In the case that the mean increases to 0.1, the algorithm is not able to locate the change point, and incorrectly determines that there is no change. In the case that the mean increases to 1, the algorithm strongly indicates that the change point is at time $t = 80.
```{r, echo = FALSE}
set.seed(8092018)
dd <- function(x,y) {
  sum((x-y)^2)^1/2
}
library(cpbaywave)
ts <- matrix(c(rnorm(80*100), rnorm(48*100,0.1)), nrow = 100, byrow = FALSE) 
dists_1 <- matrix(rep(0, 12800), nrow = 100)
for(i in 1:100) {
  coord <- sample(128, 2)
  A <- ts[,coord[1]]
  B <- ts[,coord[2]]
  den <- dd(A, B)
  for(j in 1:128) {
    dists_1[i,j] <- (dd(A, ts[,j])^2 - dd(B, ts[,j])^2)/den
  }
}
dists_1 <- t(dists_1)

ts <- matrix(c(rnorm(80*100), rnorm(48*100,1)), nrow = 100, byrow = FALSE) 
dists_2 <- matrix(rep(0, 12800), nrow = 100)
for(i in 1:100) {
  coord <- sample(128, 2)
  A <- ts[,coord[1]]
  B <- ts[,coord[2]]
  den <- dd(A, B)
  for(j in 1:128) {
    dists_2[i,j] <- (dd(A, ts[,j])^2 - dd(B, ts[,j])^2)/den
  }
}
dists_2 <- t(dists_2)


```

```{r, echo = FALSE, fig.height = 3, fig.width = 3, fig.align = "center", cache = TRUE}
cpbaywave::bootJLDetectChangePoint(dists_1)$sigindices
```

Note that in this case, the only time that is above the bootstrap significant bar is time $t = 1$. 

```{r, echo = FALSE, fig.height = 3, fig.width = 3, fig.align = "center", cache = TRUE}
cpbaywave::bootJLDetectChangePoint(dists_2)$sigindices
```

This picture gives a typical strong indication that there is a change point at or near time $t = 80$, which we know is the correct value.

We note here that the \texttt{cpbaywave} algorithm also fails to detect a change point in the first example when using the untransformed data.
\end{example}


Next, we consider an example that isn't locally Euclidean, but we apply our algorithm to it anyway.

```{r, echo = FALSE, fig.height = 3, fig.width = 3, fig.align = "center", cache = TRUE}
dd_2 <- function(x,y) {
  sum((x-y)^4)^1/4
}

ts <- matrix(c(rnorm(80*100), rnorm(48*100,1)), nrow = 100, byrow = FALSE) 
dists_3 <- matrix(rep(0, 12800), nrow = 100)
for(i in 1:100) {
  coord <- sample(128, 2)
  A <- ts[,coord[1]]
  B <- ts[,coord[2]]
  den <- dd_2(A, B)
  for(j in 1:128) {
    dists_3[i,j] <- (dd_2(A, ts[,j])^2 - dd_2(B, ts[,j])^2)/den
  }
}
dists_3 <- t(dists_3)

bootJLDetectChangePoint(dists_3)$sigindices

```

In this example, we see a drop in performance associated with the change in metric space. Nonetheless, the algorithm is still easily able to detect the change point at time $t = 80$. 

<!-- Text with citations by @Galyardt14mmm, [@Mislevy06Cog]. -->

<!-- ## Subsection title {#sec:2} -->

<!-- # Change points in persistence diagrams {#sec:2} -->



<!-- as required. Don't forget to give each section -->
<!-- and subsection a unique label (see Sect. \ref{sec:1}). -->

<!-- #### Paragraph headings  -->

<!-- Use paragraph headings as needed. -->

<!-- \begin{align} -->
<!-- a^2+b^2=c^2 -->
<!-- \end{align} -->



# Change points in persistence diagrams {#sec:2}

The space of persistence diagrams is not locally isometric to Euclidean space. However, it does seem to be closer to being isometric to Euclidean space than $\ell^4$ is. We apply our algorithm now for time series of persistence diagrams. First, we apply it to simulated data, then to data in the wild of various types.

Our first example is a proof of concept. We start with a single circle, sampled randomly, and then at time $t = 80$, we add a second disjoint circle. Any good change point detection algorithm based on persistence diagrams should be able to find such a change. Here are two plots of typical point clouds in $\mathbb{R}^2$.

```{r, echo = FALSE, fig.height = 3, fig.width = 3, fig.align = "center", cache = TRUE, message = FALSE}
library(TDA)
library(cpbaywave)
library(dplyr)
library(ggplot2)
library(boot) 
maxdimension = 1
maxscale = 5

Circles <- lapply(1:80, function(x) {
  Circle1 <- circleUnif(60);
})

Circles2 <- lapply(81:128, function(x) {
  Circle1 <- circleUnif(60);
  Circle2 <- circleUnif(60, r = 2) + 5;
  rbind(Circle1, Circle2);})


Circles <- c(Circles, Circles2)
plot(Circles[[1]])
plot(Circles[[81]])

Diags <- lapply(1:128, function(i) ripsDiag(X = Circles[[i]],
                                           maxdimension,
                                           maxscale,
                                           library = "GUDHI",
                                           printProgress = FALSE) )

dists <- matrix(1:(128*128), ncol = 128)
for(i in 1:128) {
  coord <- sample(128, 2)
  A <- Diags[[coord[1]]][[1]]
  B <- Diags[[coord[2]]][[1]]
  den <- wasserstein(A, B, p = 1)
  for(j in 1:128) {
    dists[i,j] <- (wasserstein(A, Diags[[j]][[1]], p = 1)^2 - wasserstein(B, Diags[[j]][[1]], p = 1)^2)/den
  }
}
dists <- t(dists)
bootJLDetectChangePoint(dists)$sigindices


```


Next, we consider two circles which start out being disjoint, but one of the two circles passes through the other circle. From a topological point of view, there are two potential change points. The first change point is when there are no longer two connected components, but rather one. The second change is when there are three loops rather than two. The algorithm currently under discussion only finds a single change point; how many and which dimensions are included in the persistence diagram and distance calculation will determine which change is detected. 

Here are plots from various times in the time series.

```{r, echo = FALSE, fig.height = 3, fig.width = 3, fig.align = "center", cache = TRUE}
Circles <- lapply(1:64, function(x) {
Circle1 <- circleUnif(60, r = 2) + rnorm(120,0,.15);
Circle2 <- circleUnif(100, r = 4) + 2.375 + 4*(64 - x)/64 + rnorm(200,0,.15);
#unifNose <- matrix(runif(20, -2, 10), ncol = 2)
#unifNose <- NULL
rbind(Circle1, Circle2);})

plot(Circles[[1]])
plot(Circles[[40]])
plot(Circles[[50]])

N <- 64
Diags <- lapply(1:N, function(i) ripsDiag(X = Circles[[i]],
                                           maxdimension,
                                           maxscale,
                                           library = "GUDHI",
                                           printProgress = FALSE) )

dists <- matrix(1:(N*N), ncol = N)
for(i in 1:N) {
  coord <- sample(N, 2)
  A <- Diags[[coord[1]]][[1]]
  B <- Diags[[coord[2]]][[1]]
  den <- wasserstein(A, B, p = 1)
  for(j in 1:N) {
    dists[i,j] <- (wasserstein(A, Diags[[j]][[1]], p = 1)^2 - wasserstein(B, Diags[[j]][[1]], p = 1)^2)/den
  }
}
dists <- t(dists)
bootJLDetectChangePoint(dists)$sigindices
```

<!-- The computation of Wasserstein distance between persistence diagrams can take quite some time when using zero dimensional homology, so we reduce to only taking 5 random pairs rather than 64 in the next computation. For that reason, we also do not bootstrap the significant indices, but rather we simply use Bayes Information Criteria, which we have found to be somewhat unreliable for data generated in this way. -->

```{r, echo = FALSE, fig.height = 3, fig.width = 3, fig.align = "center", cache = TRUE, eval = FALSE}
Circles <- lapply(1:64, function(x) {
Circle1 <- circleUnif(30, r = 2) + rnorm(30,0,.15);
Circle2 <- circleUnif(50, r = 4) + 2.375 + 4*(64 - x)/64 + rnorm(50,0,.15);
#unifNose <- matrix(runif(20, -2, 10), ncol = 2)
#unifNose <- NULL
rbind(Circle1, Circle2);})

N <- 64
Diags <- lapply(1:N, function(i) ripsDiag(X = Circles[[i]],
                                           maxdimension,
                                           maxscale,
                                           library = "GUDHI",
                                           printProgress = FALSE) )
N <- 5
dists <- matrix(1:(5*64), ncol = 64)
for(i in 1:N) {
  coord <- sample(64, 2)
  A <- Diags[[coord[1]]][[1]]
  B <- Diags[[coord[2]]][[1]]
  den <- wasserstein(A, B, p = 1, dimension = 0)
  for(j in 1:64) {
    dists[i,j] <- (wasserstein(A, Diags[[j]][[1]], p = 1, dimension = 0)^2 - wasserstein(B, Diags[[j]][[1]], p = 1, dimension = 0)^2)/den
  }
}
dists <- t(dists)
detectChangePoint(dists)
bootJLDetectChangePoint(dists)$sigindices
```

Next, we have two examples of time series of images. In the first example, we examine the Archive of Many Outdoor Scenes. We picked a camera that was taking pictures of Table Mountain from Bloubergstrand in South Africa. For each day, we averaged all of the pictures that were taken on that day. We then used edge detection to extract edges from the pictures, and we removed the time stamp. Finally, we randomly sampled points from the detected edges of the pictures to use as our point cloud. So, the time series consists of randomly sampled points from edges of the average of all pictures on 128 consecutive days. 

In the next, we have a series of images of a cell, which divides at time $t = 23$. We use the \texttt{cannyEdges} function in the \texttt{imager} R package to detect the edges of the cells. We then randomly sample 200 points from the detected edges at each time stamp to form our time series of point clouds. We then applied the change point detection algorithm to obtain the results below.

```{r echo = FALSE, eval = FALSE}
set.seed(07042017)
circs <- list(0)
listLen <- 64
alpha1 <- .7
startPoint <- 1
for(i in  startPoint :(startPoint + listLen - 1)) {
if(i < 10) {
im <- readPNG(paste0("/Users/speegled/Downloads/E00/Frame00", i, ".png"))
}
if(i >= 10 && i < 100) {
im <- readPNG(paste0("/Users/speegled/Downloads/E00/Frame0", i, ".png"))
}
if(i >= 100) {
im <- readPNG(paste0("/Users/speegled/Downloads/E00/Frame", i, ".png"))
}
im <- im[100:400,]
im <- as.cimg(im)
#alpha1 <- findAlpha(im, .004, alpha1)
alpha1 <- 1
im <- cannyEdges(im, alpha = alpha1)
im2 <- which(im, arr.ind = TRUE)
myRows <- sample(nrow(im2), 200)
#myRows <- 1:nrow(im2)
plot(im2[myRows, 1:2], main = as.character(i))
circs[[length(circs) + 1]] <- im2[myRows,1:2]
}
circs[[1]] <- NULL
for(i in 1:listLen)
circs[[i]] <- circs[[i]]/480
save(circs, file = "circleData.dat")
```

```{r, echo = FALSE, fig.height = 3, fig.width = 3, fig.align = "center", cache = TRUE}
load("circleData.dat")
Circles <- circs
N <- 64
Diags <- lapply(1:N, function(i) ripsDiag(X = Circles[[i]],
                                           maxdimension,
                                           maxscale = 0.1,
                                           library = "GUDHI",
                                           printProgress = FALSE) )


dists <- matrix(1:(N*N), ncol = N)
for(i in 1:N) {
  coord <- sample(N, 2)
  A <- Diags[[coord[1]]][[1]]
  B <- Diags[[coord[2]]][[1]]
  den <- wasserstein(A, B, p = 1)
  for(j in 1:N) {
    dists[i,j] <- (wasserstein(A, Diags[[j]][[1]], p = 1)^2 - wasserstein(B, Diags[[j]][[1]], p = 1)^2)/den
  }
}
dists <- t(dists)
bootJLDetectChangePoint(dists)$sigindices

```


# References
